Open-source tools in R for landscape ecology

### 2021-04-29 Comments_Reviewer.Rmd
Some last typos

**Reply**: Thank you very much for considering the manuscript for publication. 
We appreciate the reviewers comments and think they all help to improve the manuscript.
We agree with reviewer 1 that the definition of landscape ecology could be shortened given the target audience of the journal. 
Thus, the introduction now puts a stronger focus on advantages of open-source software and especially R. 
Furthermore, we added explanations for all packages that were missing in the original submission as suggested by reviewer 1.
However, we think that given recommendations which packages to use is very difficult for several reasons. 
First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences and experiences of the users.
Second, we believe that trying out several packages for similar task can actually beneficial and strengthen trust in the method.
But, by modifying Figure 1 and adding Box 1 we hope to give some more guidelines on how to organize an analysis.
We agree that the survey is not the core of the paper, nevertheless, the comments by both reviewers are very valuable and improve the conclusions we can draw from it. 
Thus, we added some explanation about the survey reach and what this means for the results of the survey. \vskip 0.5cm

@@

**Reply**: We agree that the approach of Rayfield et al. 2011 is great, but we also want to point out that it was developed for a more specific domain (i.e., connectivity), while this manuscript deals with the whole field of landscape ecology.
Thus, we believe that a similar typology for all packages included in this manuscript would be too complicated to be of actual help.
Thank you for this great idea to add boxes. 
We added a box "Where to start" to give some major steps how to approach landscape ecology in R for novice users. 
We hope this will also at least partly address the issue raised by reviewer 1 about recommendations.
We also thought about a second box for mode advanced users and even developers, but decided that the target audience of the paper are most likely novice to intermediate R users and not developers and the topic of developing own packages specifically for the landscape ecology community could maybe be a future manuscript.
Also, we added a short explanation of the tidyverse and its strength.
Lastly, we provide more information about the annotated references.
\vskip 0.5cm

@@

**Reply**: Again, thank you very much for the appreciated and valuable suggestions for improving the manuscript and we hope that the revised manuscript improved compared to the previous version with help of the reviewers. 
All changes to the manuscript are indicated by the blue color. \vskip 0.5cm

@@

**Reply**: Thank you for the general encouraging words about the manuscript. 
We agree that such a long definition of landscape ecology is not needed given the target audience of the journal. 
Thus, we shortened the introduction related to landscape ecology in general and rearranged other sections of the introduction.
Most of the introduction now introduces the advantages and strength of open-source software and in particular R. While all of the authors are enthusiastic R users, we think that a comprehensive discussion of R vs. other open-source software is outside of the scope of this paper and more important not expedient in general since other software might work better for other people (e.g., Python). \vskip 0.5cm

@@

**Reply**: All authors of the manuscript are experienced R users with many years of experience using the programming language. 
Additionally, all authors even contributed several packages to the community themselves. 
Given the different background of all authors, collectively probably most packages were used by at least one of the authors, but probably not all. 
Nevertheless, we also believe that this is not necessarily important for this review. 
Furthermore, we agree with the reviewer that more specific recommendation which package to use could be useful, however, we also think that this is very difficult. 
Reasons for this include, besides others, personal preferences of the users or analysis specific circumstances. 
Lastly, the review should be as neutral as possible without favoring one package over another because pros and cons of package differ between users and all were developed with a specific idea in mind.
But, we think that our new "Box 1: Where to start" hopefully can assist novice users with their workflow. \vskip 0.5cm

@@

**Reply**: Thank you for this valuable remark. We tried to add additional explanations to concepts which might be underexplained to especially novice users which hopefully increases the value of the manuscript. \vskip 0.5cm

@@

**Reply**: We added an explanation how these package allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts. \vskip 0.5cm

@@

**Reply**: We agree that this is problematic and we weren't even aware this is happening so thanks for pointing it out. 
We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm

@@

**Reply**: Thank you for checking the table so extensively. 
We made sure all packages should be included now. 
We also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology. \vskip 0.5cm

@@

**Reply**: We added that we aware that the survey is most likely not representative for the R community and R users due to how we tried to reach the community and also due to the questions asked in the survey. 
Since the survey is not the main idea of the manuscript though, we think that the survey still might be interesting even with these shortcomings. \vskip 0.5cm

@@

**Reply**: We agree that the sentence implied that the R community is not constantly adapting to the new developments of landscape ecology which is certainly not true. 
We reworded the sentence making clearer that indeed the community constantly improves the R programming language. 
However, we do not agree with the reviewer that "[...] the problem is that there are so many packages [..]".
Even though we acknowledge that the amount of package can be overwhelming, especially for novices users, diversity of packages is a big strength - even for similar specific tasks. 
Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures. \vskip 0.5cm

@@

**Reply**: We highly agreed that this one of the main reasons why R should be widely used and we added an explanation how the script-based character of R itself allows to share and reproduce analysis. \vskip 0.5cm

@@

**Reply**: Please see our reply to the previous comment. 
We reworded the sentence to state clearer that this is indeed one of the strength of the R community. \vskip 0.5cm

@@

**Reply**: We apologize for this. 
The manuscript was written using the rticles packages with the corresponding template for Springer journals, which somehow did not include journal names. 
We fixed the citation style including journal names. \textcolor{red}{STILL NEED TO DO THIS ACTUALLY!} \vskip 0.5cm

@@

**Reply**: We agree with the reviewer that the provided script is not used for experienced R users and targeted to novices.
Thus, we tried to improve the the documentation of the code. \vskip 0.5cm

@@

**Reply**: We strongly agree with the reviewer that this information would be very valuable to judge the representation of the survey.
However, unfortunately such questions were not included.
We also acknowledge that the survey is most likely not representative for all skill levels of *R* users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
We added and clarified those shortcomings of the survey. \vskip 0.5cm

>>>>

**Reply**: Thank you very much for considering the manuscript for publication. 
We appreciate the reviewers comments and think they all help to improve the manuscript.
We agree with reviewer 1 that the definition of landscape ecology could be shortened given the target audience of the journal. 
Thus, the introduction now puts a stronger focus on advantages of open-source software and especially R. 
Furthermore, we added explanations for all packages that were missing in the original submission as suggested by reviewer 1.
However, we think that recommendations which packages to use are very difficult for several reasons. 
First, packages have different advantages and disadvantages that differ depending on e.g., the research question, the available data, or the personal preferences and experiences of the users.
Second, we believe that trying out several packages for similar tasks can actually be beneficial and strengthen trust in the method.
But, by modifying Figure 1 and adding Box 1 we hope to give some more guidelines on how to organize an analysis.
We agree that the survey is not the core of the paper, nevertheless, the comments by both reviewers are very valuable and improve the conclusions we can draw from it. 
Thus, we added some explanation about the survey reach and what this means for the results of the survey. \vskip 0.5cm

@@

**Reply**: We agree that the approach of Rayfield et al. 2011 is great, but we also want to point out that it was developed for a more specific domain (i.e., connectivity), while this manuscript deals with the whole field of landscape ecology.
Thus, we believe that a similar typology for all packages included in this manuscript would be too complicated to be of actual help.
Thank you for the great idea to add boxes. 
We added a box "Where to start" to give some major steps how to approach landscape ecology in R for novice users. 
We hope this will also at least partly address the issue raised by reviewer 1 about recommendations.
Further, we thought about a second box for more advanced users and/or developers, but decided that the target audience of the manuscript are most likely novice to intermediate R users and not developers. 
The topic of developing own packages specifically for the landscape ecology community could be an interesting future manuscript.
Also, we added a short explanation of the tidyverse and its strength.
Lastly, we provide more information about the annotated references.
\vskip 0.5cm

@@

**Reply**: Again, thank you very much for the appreciated and valuable suggestions for improving the manuscript and we hope that the revised manuscript improved compared to the previous version with help of the reviewers. 
All changes to the manuscript are indicated by the blue color in the track change version. \vskip 0.5cm

@@

**Reply**: Thank you for the general encouraging words about the manuscript. 
We agree that such a long definition of landscape ecology is not needed given the target audience of the journal. 
Thus, we shortened the introduction related to landscape ecology in general and rearranged other sections of the introduction.
Most of the introduction now introduces the advantages and strength of open-source software and in particular R. While all of the authors are enthusiastic R users, we think that a comprehensive discussion of R vs. other open-source software is outside of the scope of this paper and more importantly not expedient in general since other software might work better for other people (e.g., Python). \vskip 0.5cm

@@

**Reply**: All authors of the manuscript are experienced R users with many years of experience using the programming language. 
Additionally, all authors even contributed several packages to the community themselves. 
Given the different background of all authors, collectively probably most packages were used by at least one of the authors, but probably not all. 
Nevertheless, we also believe that this is not necessarily important for this review. 
Furthermore, we agree with the reviewer that more specific recommendation which package to use could be useful, however, we also think that this is very difficult. 
Reasons for this include, besides others, personal preferences of the users or analysis specific circumstances. 
Lastly, we believe that the review should be as neutral as possible without favoring one package over another because pros and cons of packages differ between users and all were developed with a specific idea in mind.
But, we think that our new "Box 1: Where to start" hopefully can assist novice users with their workflow. \vskip 0.5cm

@@

**Reply**: Thank you for this valuable remark. We tried to add additional explanations to concepts which might be underexplained especially for novice users. 
We hope that this increases the value of the manuscript. \vskip 0.5cm

@@

**Reply**: We added an explanation how these packages allow to incorporate functionality currently only available in GIS software into reproducible and shareable R scripts. \vskip 0.5cm

@@

**Reply**: We agree that this is problematic and we were not aware this is happening so thanks for pointing it out. 
We deactivated the automatic hyphenation by LaTeX. \vskip 0.5cm

@@

**Reply**: Thank you for checking the table so extensively. 
We made sure all packages related to spatial data and landscape ecology should be included now. 
Thus, we also added the information to the table caption that not all packages from the manuscript are included but only packages specifically designed for spatial data and landscape ecology. \vskip 0.5cm

@@

**Reply**: We added that we aware that the survey is most likely not representative for the R community due to how we tried to reach the community and also due to the questions asked in the survey. 
Since the survey is not the main idea of the manuscript though, we think that the survey still might be interesting - even with these shortcomings. \vskip 0.5cm

@@

**Reply**: We agree that the sentence implied that the R community is not constantly adapting to the new developments of landscape ecology which is certainly not true. 
We reworded the sentence making clearer that indeed the community constantly improves the R programming language. 
However, we do not agree with the reviewer that "[...] the problem is that there are so many packages [..]".
Even though we acknowledge that the amount of package can be overwhelming, especially for novices users, diversity of packages is a big strength - even for similar specific tasks. 
Possible advantages include e.g., the possibility to compare different implementations of similar/identical task or providing functionality for slightly different data structures and user preferences. \vskip 0.5cm

@@

**Reply**: We highly agreed that this one of the main reasons why R should be widely used and we added an explanation how the script-based characteristic of R itself allows to share and reproduce analysis. \vskip 0.5cm

@@

**Reply**: We reworded the sentence to state clearer that this is indeed one of the strength of the R community (Please see also our reply to the previous comment regarding this). \vskip 0.5cm

@@

**Reply**: We apologize for this.
There was an issue with our bibliography file, which we fixed and the citation style includes journal names now. \vskip 0.5cm

@@

**Reply**: We agree with the reviewer that the provided script is not useful for experienced R users and targeted to novices users.
Thus, we tried to improve the the documentation of the code. \vskip 0.5cm

@@

**Reply**: We strongly agree with the reviewer that this information would be very valuable to judge the representation of the survey.
However, unfortunately such questions were not included.
We also acknowledge that the survey is most likely not representative for all skill levels of R users because we assume that by using mailing list and social media more advanced users were more likely to take part in the survey.
We added and clarified those shortcomings of the survey. \vskip 0.5cm

### 2021-04-29 paper/paper.Rmd
A few small changes.
Some last typos

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as, those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: sentence
---

@@

While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data. 
\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.} 
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, page 10, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
Because all \textit{tidyverse} packages follow the same data philosophy structures, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at many scales is available from an abundance of online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the `getData()` function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for CRS storing.
Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}

@@

\textbf{4.} Do a quick visualization of the data for inspection or a first glance at the data. 
This can be done with base \textit{R} or if already familiar - \textit{ggplot2}. \\

@@

\multirow{3}{*}{Bridge stuff} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\

@@

\multirow{5}{*}{Connectivity} & \textit{lconnect} & Calculate landscape connectivity metrics & \cite{Mestre2019} \\

@@

To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community. \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey.}.
All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity", and "nature conservation" (Fig. \@ref(fig:fig-survey)a)).

@@

Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated \textit{R} as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig. \@ref(fig:fig-survey)d)).

>>>>

---
title: Open-source tools in R for landscape ecology
authors: 
- name: Maximillian H.K. Hesselbarth
  address:
    a) Department of Ecosystem Modelling, University of Goettingen, Buesgenweg 4, 37077 Goettingen, Germany
    b) Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, MI, 48109, USA
  email: mhk.hesselbarth@gmail.com
- name: Jakub Nowosad
  address: Institute of Geoecology and Geoinformation, Adam Mickiewicz University, Krygowskiego 10, 61-680 Poznan, Poland
  email: nowosad.jakub@gmail.com
- name: Johannes Signer
  address: Wildlife Sciences, Faculty of Forestry and Forest Ecology, University of Goettingen, Buesgenweg 3, 37077 Goettingen, Germany
  email: jsigner@uni-goettingen.de
- name: Laura J. Graham
  address: 
    a) Geography, Earth & Environmental Sciences, University of Birmingham, Edgbaston, Birmingham B15 2TT, United Kingdom
    b) Biodiversity, Ecology & Conservation Group, IIASA, Vienna, Austria 
  email: l.graham@bham.ac.uk
keywords:
- spatial data
- statistical programming language
- "*R* packages"
- reproducibility
- scientific software
abstract: |
  \textbf{Purpose of review} Landscape ecology, the study of the complex interactions between landscapes and ecological processes, has hugely benefited from the increase in widely available open source software in recent years. In particular, the \textit{R} programming language provides a wealth of community developed tools for landscape ecology. \textbf{Recent findings} In this paper, we examine existing packages for downloading, processing and visualisation of spatial data, as well as, those specifically developed for ecological analysis. Additionally, we outline the results of a survey of \textit{R} users within the landscape ecology community. \textbf{Summary} We found that landscape ecologists are generally satisfied with the functionality available within \textit{R}, and that as a community they are continually further develop the functionality available. Suggested future developments include improvement of computation performance; additional methods for landscape characterisation such as surface metrics; and advanced, accessible visualisation tools. 
bibliography: bibliography.bib
bibstyle: spphys # bibstyle options spbasic(default), spphys, spmpsci
preamble: |
  \usepackage{caption}
  \usepackage{tabularx}
  \usepackage{multirow}
  \usepackage{placeins}
  \usepackage{float}
  \usepackage[modulo]{lineno}
  \usepackage{xcolor}
  \usepackage[none]{hyphenat}
output:
  bookdown::pdf_book:
    base_format: rticles::springer_article
    number_sections: yes
    citation_package: default
editor_options:
  chunk_output_type: console
---

@@

While \textcolor{blue}{base \textit{R}} has several built-in data structures, including vectors, matrices, data frames, and lists, it has no internal support for reading, processing, or visualizing spatial data. 
\textcolor{blue}{However, as discussed previously, one strength and core idea of the \textit{R} programming language are its expandability by packages.} 
Because there is a substantial interest in spatial data analysis, support for spatial data is now provided by many \textit{R} packages (\cite{Lovelace2019}, Table \@ref(tab:packages)).
Most spatial data belong to one of two data models, namely spatial raster and spatial vector model, and both data models have several implementations in the \textit{R} language.
Importantly, main \textit{R} packages for spatial data use the external \textit{GDAL} \cite{GDAL/OGRcontributors2020} and \textit{PROJ} \cite{PROJcontributors2021} libraries, which allow for reading and writing of hundreds of spatial data formats, and for coordinates transformation.
Additionally, \textit{R} allows for conversion between data models and specific implementations, which can be useful if given methods only exist for a particular data model or implementation.

@@

The vector data model consists of two main elements i) geometries (such as points, lines, polygons) and ii) attributes, where each geometry is connected to a row in the attribute table.
In many cases, this data model allows a more realistic representation of landscape features, however, with the higher computational cost \cite{Lovelace2019}.
The \textit{sp} package was the standard for vector data representation for more than ten years \cite{Pebesma2005,Bivand2013}.
As of 2020, more than 500 \textit{R} packages directly depend or imports \textit{sp}.
However, \textit{sp} is not actively developed anymore, and its recommended successor is the \textit{sf} package \cite{Pebesma2018}.
Besides many advantages and strengths of the \textit{sf} package in terms of spatial data handling, it also integrates into the widely used \textit{tidyverse} packages \cite{Wickham2019}.
\textcolor{blue}{The \textit{tidyverse} is a collection of \textit{R} packages developed for almost all major tasks of any data analysis project. 
Because all \textit{tidyverse} packages follow the same philosophy how to structure data, one strength of the \textit{tidyverse} is its high consistency of usage across its packages \cite{Wickham2019}.}
\textit{sf} builds on the idea of "simple features", a standard used to describe spatial geometries using points, lines (two connected points) and polygons (several connected points), and attributes connected to these geometries \cite{Pebesma2019a}.

@@

Nowadays, spatial data at various scales is available from many online-accessible sources.
A lot of this data are publicly available, either as a direct download or through an API connection, and several packages can use this to download the spatial data directly into an \textit{R} session.
Since publicly available data is becoming more prominent, so are \textit{R} packages to access them.
Packages include \textit{rnaturalearth} \cite{South2017} to access the Natural Earth database to download region and country data, the \textit{elevatr} package to access raster elevation data \cite{Hollister2020}, the \textit{rgbif} package to access the Global Biodiversity Information Facility (GBIF) portal \cite{Chamberlain2017}, the \textit{BIEN} package \cite{Maitner2020} to access the Botanical Information and Ecology Network Database, the \textit{marmap} to download bathymetry data from the ETOPO1 database \cite{Pante2013}, or the \textit{FedData} package \cite{Bocinsky2019} to access the National Land Cover Database (NLCD) data for the USA.
Furthermore, the \textit{getlandsat} package \cite{Chamberlain2018} allows users to download Landsat 8 satellite data, the \textit{MODIS} package \cite{Mattiuzzi2020} to download MODIS products, and \textit{sen2r} \cite{Ranghetti2020} to download Sentinel-2 optical images.
Also, the `getData()` function from the \textit{raster} package allows users to download climatic and bioclimatic data from WorldClim v1.4.
Additionally, the \textit{rgee} package \cite{Aybar2020} gives access to an extensive catalog of data from Google Earth Engine, including climate data, land cover maps, and satellite imagery.

@@

\textcolor{blue}{Until 2020, spatial \textit{R} objects stored information about their CRS using Proj4 string representations, and these strings could also be used for CRS transformations.
However, the major changes in upstream software, namely PROJ6 and GDAL3, limited the capabilities of the Proj4 strings and switched to use a new representation called WKT2-2019 for storing the CRS.
Thus, spatial objects saved with degraded Proj4 string representation to \textit{R} file formats could be incorrectly read, and special attention should be given to check their correctness.
Proj4 string representation is also not recommended to set or transform CRS and instead the previously introduced EPSG codes should be used.
A more detailed explanation of the recent CRS changes can be found in \cite{Bivand2020}.}

@@

\textbf{4.} Do a quick visualization of the data for inspection or a first glance at the data. 
This can be done with base \textit{R} or -- if already familiar -- \textit{ggplot2}. \\

@@

\multirow{3}{*}{Interfacing to GIS} & rgrass7 & Interface between GRASS7 and R  & \cite{Bivand2021} \\

@@

\multirow{5}{*}{Connectivity} & lconnect & Calculate landscape connectivity metrics & \cite{Mestre2019} \\

@@

To better understand how the landscape ecology community uses \textit{R}, we conducted a short survey using mailing lists and social media to reach the community. \textcolor{blue}{Thus, the survey was not necessarily representative for all skill levels of \textit{R} users and results are most likely biased towards more advanced users.
Also, the survey did not include any personal questions, such as age or country of residence, which might affect the representation of the survey}.
All raw data and scripts to analyze the data can be found at [https://github.com/r-spatialecology/Hesselbarth_et_al_CLER](https://github.com/r-spatialecology/Hesselbarth_et_al_CLER).

@@

We asked the participants to select which terms describe their research topics the best, and options that were selected by more than 10% of participants included "biodiversity", followed by "land use management", "landscape connectivity", and "nature conservation" (Fig. \@ref(fig:fig-survey)a).

@@

Lastly, when asked how useful \textit{R} is currently for landscape ecology, the vast majority of participants answered with either "very useful" or "useful" (summarized `r good_eval`%) and only very few participants evaluated \textit{R} as "intermediate", "not useful" or "not useful at all" (summarized `r bad_eval`%; Fig. \@ref(fig:fig-survey)d).
